---
title: Working with data frames
description: |
  Basic principles to work with data frame in base R, Tidyverse, and data.table.
author: R.Andres Castaneda
date: '2023-01-31'
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
xaringanExtra::use_panelset()
```

# Set up

Attach important packages. For a comprehensive comparison see this [blog](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/). 

```{r}
library(tidyverse)
library(data.table)
```

Load data. You can go [here](https://github.com/PovcalNet-Team/Rtraining/raw/main/data) to take a look at some fake data in different formats

```{r, results=FALSE}

link_data <- "https://github.com/PovcalNet-Team/Rtraining/raw/main/data/ago_2018.csv"

df <- read.csv(link_data) # base
tb <- read_csv(link_data)  # tidyverse 
dt <- fread(link_data) # data.table

```


We could have done also this

```{r eval=FALSE}
df <- read.csv(link_data) # base
tb <- as.tibble(df)
dt <- as.data.table(tb)
```


# Basic operations

## Filter rows

### Keep rows using indices
```{r}
filter <- c(3:4)
```

::: panelset
::: panel

#### Base R

```{r, error=TRUE}
df[filter,]
df[filter] # This does not work
```
:::

::: panel
#### Tidyverse

```{r}
tb[filter,]
slice(tb, filter) # same
```
:::

::: panel
#### data.table

```{r}
dt[filter,]
# This works. In data.frame does not. 
dt[filter] # same. 
```
:::
:::

<aside>For the sake of time, I'll incorporate all the examples later.</aside>

### Keep rows using logical expressions

::: panelset
::: panel

#### Base R

```{r}
x <- df[df$area == "urban",]
x[1:3,] 
```
:::

::: panel
#### Tidyverse

```{r}
tb |> 
  filter(area == "urban") |> 
  slice(1:3)
```
:::

::: panel
#### data.table

```{r, error=TRUE}

# data.table way and No need of $
 
dt[area == "urban"
   ][1:3]

# Tidyverse syntax works with data.table
dt |> 
  filter(area == "urban") |> 
  slice(1:3)

# but data.table syntax does not with tidyverse
tb[area == "urban"][1:3]

```
:::
:::


