---
title: Preliminary notions
description: |
  Básic concepts such as object, names, types, and class.
author: R.Andres Castaneda
date: '2022-11-22'
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Slides


```{r embed-class01, echo=FALSE}
xaringanExtra::embed_xaringan(url   = "../../slides/preliminary_notions/",
                              ratio = "16:9")
```

#  Hands on 

## Exercises 


1. What is the difference between `c(list(1:3), list(9))` and `list(list(1:3), list(9))`

2. Is there any difference between these three ways of assignment?

```{r ex2}
assign('a', c(10, 150, 30, 45, 20.3))
a <- c(10, 150, 30, 45, 20.3)
a = c(10, 150, 30, 45, 20.3)
```

3. The `dim()` function gets the dimension of an object. Why `dim(a)` is `NULL`?

4. Why `is.character(c('blue',10,'green',20))` is `TRUE`?

R comes with a variety of datasets to work with (Similar to `sysuse auto` in Stata). You can see all the datasets available by typing `library(help = "datasets")`.  

5. In a `tibble` bound to the name `tb`, add a column named `x` with the vector `rivers` from the internal `datasets`.



## Understanding factors

### From characters to factors
```{r chr}

## Simple character vector
degree <- c("master", "master", "bachelor", "phd", "bachelor", "master")
degree
class(degree)
typeof(degree)
attributes(degree)
summary(degree)

```


```{r fc}

degree_fc <- factor(degree, levels = c("bachelor", "master",  "phd"))
class(degree_fc)
typeof(degree_fc)
attributes(degree_fc)
summary(degree_fc)
```

```{r fc-plot}
plot(degree_fc, col = "steelblue")

```

```{r fc-plot-err, error=TRUE}
plot(degree, col = "steelblue")

```

### From integer to factors

```{r fc-int}

degree2    <- c(1L, 3L, 1L)
class(degree2)
typeof(degree2)
attributes(degree2)
summary(degree2)

degree2_fc <- factor(degree2,
                    levels = c(1, 2, 3),
                    labels = c("bachelor", "master", "phd"))
class(degree2_fc)
typeof(degree2_fc)
attributes(degree2_fc)
summary(degree2_fc)


plot(degree2)
plot(degree2_fc)

```


## The beauty of classes


```{r methods}

methods("summary")

methods("print")

```

