---
title: How to use the Pipe
description: |
  Understand the difference between the magritr pipe and the native pipe, and when 
  to use each. Understand when it is better to use the pipe rather than regular 
  Base R syntax. 
author: R.Andres Castaneda
date: '2023-05-09'
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Packages

## loading and attaching

There is an important difference between loading and attaching a package.

-   Loading refers to put all the components of a package available in memory ( code, data, and any DLLs; register S3 and S4 methods). However, those components are not in the search the **search path**, which is equivalent to the **ado path** in Stata. This is way, we need to call the function of loaded package with `::`. If the packages has not been loaded, using `::` loads the package.

-   Attaching loads the package and makes it available in the search path. You do it using `library()` or `require()`. When it is attached, you don't need to use `::`, but you can.

Read more [here](https://r-pkgs.org/dependencies-mindset-background.html#sec-dependencies-attach-vs-load), R-Packages book.

## When to attach

Attaching has the advantage of not using `::`, but when you use too many packages, it is difficult to know what function comes from what package.

I attach when I am using a package that I need all the time. For example, `tidyverse` or `data.table`. However, try to always use `::` because the code is clearer and there is no penalty in speed (minimum).

If you are developing packages, you canNOT attach package. You always have to use `::`

# Main idea of the pipe

```{r}
# 
```

# sources

-   This post in [stackoverflow](https://stackoverflow.com/a/72086492/11472481)
-   This [blog](https://ivelasq.rbind.io/blog/understanding-the-r-pipe/) by Isabella Velásquez. It is outdated because it is based in R 4.1, but it is still useful.
