---
title: How to use the Pipe
description: |
  Understand the difference between the magritr pipe and the native pipe, and when 
  to use each. Understand when it is better to use the pipe rather than regular 
  Base R syntax. 
author: R.Andres Castaneda
date: '2023-05-09'
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Packages

## loading and attaching

There is an important difference between loading and attaching a package.

-   Loading refers to put all the components of a package available in memory ( code, data, and any DLLs; register S3 and S4 methods). However, those components are not in the search the **search path**, which is equivalent to the **ado path** in Stata. This is way, we need to call the function of loaded package with `::`. If the packages has not been loaded, using `::` loads the package.

-   Attaching loads the package and makes it available in the search path. You do it using `library()` or `require()`. When it is attached, you don't need to use `::`, but you can.

Read more [here](https://r-pkgs.org/dependencies-mindset-background.html#sec-dependencies-attach-vs-load), R-Packages book.

## When to attach

Attaching has the advantage of not using `::`, but when you use too many packages, it is difficult to know what function comes from what package.

I attach when I am using a package that I need all the time. For example, `tidyverse` or `data.table`. However, try to always use `::` because the code is clearer and there is no penalty in speed (minimum).

```{r, eval=FALSE}
flights |> 
  dplyr::filter(dest == "IAH") |> 
  dplyr::mutate(speed = distance / air_time * 60) |> 
  dplyr::select(year:day, dep_time, carrier, flight, speed) |> 
  dplyr::arrange(dplyr::desc(speed))


flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))

```

If you are developing packages, you **cannot** attach packages. You always have to use `::`

# Main idea of the pipe

```{r}
# load libraries that we will need

library(nycflights13) # data or use library(help = "datasets")
library(tidyverse)
library(data.table)

```

We need to talk first about frames. Let's go to Stata first.

At the most basic level, the pipe is a syntax transformation in which you separate the argument from the function

```{r}
x = 1:10

# from this
mean(x)

# to this 

x |> mean()
```

But it by itself is not super useful. You see the real power when you work with dataframes

```{r, eval=FALSE}
# so you go from this
flights1 <- filter(flights, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))


# or this
arrange(
  select(
    mutate(
      filter(flights, dest == "IAH"),
      speed = distance / air_time * 60
    ),
    year:day, dep_time, carrier, flight, speed
  ),
  desc(speed)
)


# to this


```

To this.

```{r}
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```

# sources

-   This post in [stackoverflow](https://stackoverflow.com/a/72086492/11472481)
-   This [blog](https://ivelasq.rbind.io/blog/understanding-the-r-pipe/) by Isabella Velásquez. It is outdated because it is based in R 4.1, but it is still useful.
