---
title: Subsetting
description: |
  select elements from an R object. All the informtion in this 
  chapter comes from the Subsetting chapter of Advanced R by 
  Hadley Wickham.
author: R.Andres Castaneda
date: '2023-01-24'
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Atomic vectors

We have two vectors, `x` and `y` .

```{r}
x <- c(1, 3, 2, 4, -10)
y <- c("c", "a", "d",  "b")
length(x)
length(y)
```

Use [`[`](https://rdrr.io/r/base/Extract.html) to select any number of elements from a vector.

## Using numbers

### Positive numbers

Return the elements in **specified positions**.

```{r}
x[c(3, 1)]

# Order of the elements of `x` is another vector of the same length as `x`
order(x)
order(y)

x
y

x[order(x)]
x[order(y)]
y[order(x)]

# Duplicate indices will duplicate values
y
y[c(1, 1)]

# Real numbers are silently truncated to integers
y
y[c(2.1, 2.9)]

```

### Negative numbers

Exclude elements at the specified positions:

```{r}
# both element ar negative
x[-c(3, 1)]
```

you can't mix positive and negative integers in a single subset

```{r , error=TRUE}
x[c(-1, 2)]
```

But you can do the following

```{r}
c(x[-1], x[2])
```

### Logical vectors

Logical vectors are recycled. For subsetting `c(1, 0)` is different to `c(TRUE, FALSE)`.

```{r}
# For subsetting
y
y[c(FALSE, TRUE, TRUE, FALSE)]
y[c(0, 1, 1, 0)]

# Length zero
y[0]

# Return original vector
y[]

# Logical evaluation 

if (TRUE) {
  print("works")
}
if (1) {
  print("works")
}

if (FALSE) {
  # Not working
  print("does not work")
}
if (0) {
  # Not working
  print("does not work")
}

# Rescycle of logical vectors for subsetting
x
x[c(TRUE, FALSE)]
```

## Using names

Sometimes, elements of a vector are named. Remember, this is *different* from having a vector named. The variables in a data frame are **named** vectors, but you could also name the elements of a variable. This is similar, but not the same to have a factor vector. Factors are numeric vectors with a class `factor`. Named elements in a vector are elements of any `type` with a corresponding name. Factors behave like factors depending of the function that is reading them. Named vectors behave according to their own `type`, regardless of the name. Yes, class and names are attributes but of different kinds.

```{r}

# using setNamnes()
nombres <- c("Serapio", "Trimegisto", "Amalasunta", "Metafrasto", "Brunilda")
xm <- setNames(object = x, 
               nm = nombres)
xm

# using names()
x
names(x) <- nombres
x

# Print or use the names
names(x)

nm <- names(x)
nm
# Remove names
names(x) <- NULL
x


# Getting one element
xm["Amalasunta"]

# getting an element from a vector without names

x["Amalasunta"]

# Some names exist and some other not
xm[c("Andres", "Amalasunta")]

# repeated names
xm[c("Trimegisto", "Trimegisto")]


# Names are matched exaxtly.
xm[c("Tri", "Trimegisto")]

```

### Subsetting with factors

It is just a bad idea.

```{r}
fnames <- factor(nombres)
fnames
as.numeric(fnames)
fnames <- factor(nombres)

# factors are not names
f2 <- fnames[c("Metafrasto", "Serapio")]
f2

f3 <- fnames[c(3, 5)]

xm[c(3, 5)]
xm[f3]

as.numeric(f3)

```

# Exercises
